Cleaning script

### 2025-04-18 code/analysis-code/statistical-analysis.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

```{r}
#load needed packages. make sure they are installed.
# Load required packages. Ensure they are installed.
library(survival)  # For Cox regression analysis
library(survminer) # For survival curve visualization
library(broom)     # For tidying model output
library(tidyverse) # For data manipulation
library(here)      # For file path management (ensures reproducibility)
library(stringr)  # Ensure string manipulation functions are available
library(tidymodels)
library(vip)
library(themis)
library(randomForestSRC)
library(dplyr)
library(ranger)
library(survival)
library(pROC)
library(caret)
library(glmnet)
```

@@

@@

@@

# Rename terms for better readability (comparing each group to 20-24 years)
cox_age_results <- cox_age_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    str_detect(term, "age_group25-29 years") ~ "25-29",
    str_detect(term, "age_group30-34 years") ~ "30-34",
    str_detect(term, "age_group35-39 years") ~ "35-39",
    str_detect(term, "age_group40-44 years") ~ "40-44",
    str_detect(term, "age_group45-49 years") ~ "45-49",
    str_detect(term, "age_group50-54 years") ~ "50-54",
    str_detect(term, "age_group55-59 years") ~ "55-59",
    str_detect(term, "age_group60-64 years") ~ "60-64",
    str_detect(term, "age_group65-69 years") ~ "65-69",
    str_detect(term, "age_group70-74 years") ~ "70-74",
    str_detect(term, "age_group75-79 years") ~ "75-79",
    str_detect(term, "age_group80-84 years") ~ "80-84",
    str_detect(term, "age_group85+ years") ~ "85+",
    TRUE ~ term  # Keep other terms unchanged
  ))

@@

@@

# Optional: Save selected variables
write.csv(lasso_selected, "results/tables/lasso_selected_variables.csv", row.names = FALSE)

>>>>

```{r}
#load needed packages. make sure they are installed.
# Load required packages. Ensure they are installed.
library(survival)  # For Cox regression analysis
library(survminer) # For survival curve visualization
library(broom)     # For tidying model output
library(tidyverse) # For data manipulation
library(here)      # For file path management (ensures reproducibility)
library(stringr)  # Ensure string manipulation functions are available
library(tidymodels)
library(vip)
library(themis)
library(randomForestSRC)
library(dplyr)
library(ranger)
library(survival)
library(pROC)
library(caret)
library(glmnet)
library(gt)
library(dplyr)
library(here)
library(webshot2)
library(fs)
```

@@

```
```{r}
# Publication Quality table
# Step 1: Create survival object (if not already created)
# Replace with your actual survival time and event columns
# surv_obj <- Surv(time = mydata_filtered$survival_months, event = mydata_filtered$death_flag)

# Step 2: Run Cox regression function with conf.int = TRUE
run_cox <- function(var) {
  formula <- as.formula(paste("surv_obj ~", paste0("`", var, "`")))
  cox_model <- coxph(formula, data = mydata_filtered)
  result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)
  return(result)
}

# Step 3: Run the function for each metastasis site
cox_bone  <- run_cox("mets_bone")
cox_brain <- run_cox("mets_brain")
cox_liver <- run_cox("mets_liver")
cox_lung  <- run_cox("mets_lung")

# Step 4: Combine and relabel terms
cox_results <- bind_rows(cox_bone, cox_brain, cox_liver, cox_lung) %>%
  mutate(term = case_when(
    term == "mets_boneYes"  ~ "Bone Metastasis",
    term == "mets_brainYes" ~ "Brain Metastasis",
    term == "mets_liverYes" ~ "Liver Metastasis",
    term == "mets_lungYes"  ~ "Lung Metastasis",
    TRUE ~ term
  ))

# Step 5: Format for presentation
cox_formatted <- cox_results %>%
  mutate(
    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
    significant = ifelse(p.value < 0.05, TRUE, FALSE)
  ) %>%
  select(Metastasis = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, significant)

# Step 6: Create GT table with red highlight for significant rows
# Format for presentation but do not include `significant` column in final output
cox_formatted <- cox_results %>%
  mutate(
    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
    highlight = ifelse(p.value < 0.05, TRUE, FALSE)  # internal use only
  )

# GT table without the highlight column visible
mets_cox_gt <- cox_formatted %>%
  select(Metastasis = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, highlight) %>%
  gt() %>%
  tab_header(
    title = md("**Table 5: Association Between Metastasis and Overall Survival**")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_text(color = "red"),
    locations = cells_body(
      columns = c(`Hazard Ratio (95% CI)`),
      rows = highlight == TRUE
    )
  ) %>%
  opt_row_striping() %>%
  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    data_row.padding = px(4),
    table.align = "center",
    table_body.hlines.color = "gray90",
    table.border.top.color = "gray80",
    table.border.bottom.color = "gray80"
  ) %>%
  tab_source_note(md("*Note:* Statistically significant hazard ratios (p < 0.05) are shown in red.")) %>%
  cols_hide(columns = c(highlight))  # 👈 Hides the internal logic column

# Step 7: Save outputs
saveRDS(cox_results, here("results", "tables", "mets_HR.rds"))
saveRDS(mets_cox_gt, here("results", "tables", "mets_HR_gt.rds"))

dir_create(here("results", "figures"))

gtsave(mets_cox_gt,
       filename = here("results", "figures", "mets_HR_table.png"),
       vwidth = 1000, vheight = 600, zoom = 2)
#View table here
mets_cox_gt

@@

```
```{r}
#Publication-ready Table 
# Step 1: Recode the race variable (if not already done)
mydata_filtered <- mydata_filtered %>%
  mutate(race = factor(race, levels = c(
    "White", "Black", "Asian or Pacific Islander", "American Indian/Alaska Native"
  )))

# Step 2: Define the Cox model function (includes CI)
run_cox_race <- function() {
  formula <- as.formula("surv_obj ~ race")
  cox_model <- coxph(formula, data = mydata_filtered)
  result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)  # ✅ Includes CI
  return(result)
}

# Step 3: Run the model
cox_race_results <- run_cox_race()

# Step 4: Rename comparisons
cox_race_results <- cox_race_results %>%
  mutate(term = case_when(
    term == "raceBlack" ~ "Black vs. White",
    term == "raceAsian or Pacific Islander" ~ "Asian vs. White",
    term == "raceAmerican Indian/Alaska Native" ~ "American Indian vs. White",
    TRUE ~ term
  ))

# Step 5: Format HR (95% CI) and significance
cox_race_formatted <- cox_race_results %>%
  mutate(
    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
    highlight = ifelse(p.value < 0.05, TRUE, FALSE)
  )

# Step 6: Create GT table with red highlights
race_cox_gt <- cox_race_formatted %>%
  select(Comparison = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, highlight) %>%
  gt() %>%
  tab_header(
    title = md("**Table 6: Association Between Race and Overall Survival**")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_text(color = "red"),
    locations = cells_body(
      columns = c(`Hazard Ratio (95% CI)`),
      rows = highlight == TRUE
    )
  ) %>%
  opt_row_striping() %>%
  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    data_row.padding = px(4),
    table.align = "center",
    table_body.hlines.color = "gray90",
    table.border.top.color = "gray80",
    table.border.bottom.color = "gray80"
  ) %>%
  tab_source_note(md("*Note:* Statistically significant hazard ratios (p < 0.05) are shown in red.")) %>%
  cols_hide(columns = c(highlight))

# Step 7: Show the table in QMD output
race_cox_gt

# Step 8: Save outputs
saveRDS(cox_race_formatted, here("results", "tables", "race_HR.rds"))
saveRDS(race_cox_gt, here("results", "tables", "race_HR_gt.rds"))

dir_create(here("results", "figures"))
gtsave(race_cox_gt,
       filename = here("results", "figures", "race_HR_table.png"),
       vwidth = 1000, vheight = 600, zoom = 2)

# Show table in HTML/preview
race_cox_gt

@@

# Rename terms for better readability (comparing each group to 20-24 years)
cox_age_results <- cox_age_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    str_detect(term, "age_group25-29 years") ~ "25-29",
    str_detect(term, "age_group30-34 years") ~ "30-34",
    str_detect(term, "age_group35-39 years") ~ "35-39",
    str_detect(term, "age_group40-44 years") ~ "40-44",
    str_detect(term, "age_group45-49 years") ~ "45-49",
    str_detect(term, "age_group50-54 years") ~ "50-54",
    str_detect(term, "age_group55-59 years") ~ "55-59",
    str_detect(term, "age_group60-64 years") ~ "60-64",
    str_detect(term, "age_group65-69 years") ~ "65-69",
    str_detect(term, "age_group70-74 years") ~ "70-74",
    str_detect(term, "age_group75-79 years") ~ "75-79",
    str_detect(term, "age_group80-84 years") ~ "80-84",
    str_detect(term, "age_group85\\+ years") ~ "85+",
    TRUE ~ term  # Keep other terms unchanged
  ))

@@

```
```{r}
#Publication-ready table

# Step 1: Re-run the Cox model with confidence intervals
run_cox_age <- function() {
  formula <- as.formula("surv_obj ~ age_group")
  cox_model <- coxph(formula, data = mydata_filtered)
  result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)  # <-- CI INCLUDED
  return(result)
}

cox_age_results <- run_cox_age()

# Step 2: Relabel age group comparisons (optional cleaning)
cox_age_results <- cox_age_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    str_detect(term, "age_group25-29 years") ~ "25–29",
    str_detect(term, "age_group30-34 years") ~ "30–34",
    str_detect(term, "age_group35-39 years") ~ "35–39",
    str_detect(term, "age_group40-44 years") ~ "40–44",
    str_detect(term, "age_group45-49 years") ~ "45–49",
    str_detect(term, "age_group50-54 years") ~ "50–54",
    str_detect(term, "age_group55-59 years") ~ "55–59",
    str_detect(term, "age_group60-64 years") ~ "60–64",
    str_detect(term, "age_group65-69 years") ~ "65–69",
    str_detect(term, "age_group70-74 years") ~ "70–74",
    str_detect(term, "age_group75-79 years") ~ "75–79",
    str_detect(term, "age_group80-84 years") ~ "80–84",
    str_detect(term, "age_group85\\+ years") ~ "85+",
    TRUE ~ term
  ))

# Step 3: Format table with HR, CI, p-value, significance flag
cox_age_formatted <- cox_age_results %>%
  mutate(
    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
    highlight = ifelse(p.value < 0.05, TRUE, FALSE)
  )

# Step 4: Create GT table
age_cox_gt <- cox_age_formatted %>%
  select(`Age Group (vs. 20–24)` = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, highlight) %>%
  gt() %>%
  tab_header(
    title = md("**Table 6: Association Between Age Group and Overall Survival**")
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_text(color = "red"),
    locations = cells_body(
      columns = c(`Hazard Ratio (95% CI)`),
      rows = highlight == TRUE
    )
  ) %>%
  opt_row_striping() %>%
  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    data_row.padding = px(4),
    table.align = "center",
    table_body.hlines.color = "gray90",
    table.border.top.color = "gray80",
    table.border.bottom.color = "gray80"
  ) %>%
  tab_source_note(md("*Note:* Statistically significant hazard ratios (p < 0.05) are shown in red.")) %>%
  cols_hide(columns = c(highlight))

# Step 5: Save GT object and PNG figure
saveRDS(age_cox_gt, here("results", "tables", "age_HR_gt.rds"))

dir_create(here("results", "figures"))

gtsave(age_cox_gt,
       filename = here("results", "figures", "age_HR_table.png"),
       vwidth = 1000, vheight = 600, zoom = 2)

# Optional: display in output
age_cox_gt
```

@@

saveRDS(lasso_selected, file = here::here("results", "tables", "lasso_selected.rds"))
saveRDS(cvfit, file = here::here("results", "tables", "cvfit_lasso.rds"))
```

### 2025-04-18 code/eda-code/edamz.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

@@

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "edatables", "edasummarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```
```{r}
# Create a structured summary table
var_summary <- mydata %>%
  summarise_all(class) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Type") %>%
  mutate(Unique_Values = sapply(mydata, function(x) length(unique(x))),
         Missing_Values = sapply(mydata, function(x) sum(is.na(x))))

@@

# Save the structured table in a readable format
summarytable_file = here("results", "edatables", "edasummarytable2.rds")
saveRDS(var_summary, file = summarytable_file)
```

@@

```
Now lets look at the bar plot for age distribution

@@

>>>>

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(gt)
library(webshot2)
library(fs)
library(dplyr)
```

@@

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "edatables", "edasummarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```
```{r}
# Publication ready-table
# Get skim summary
summary_df <- skim(mydata)

# Filter only for survival_months (numeric type)
surv_summary <- summary_df %>%
  filter(skim_variable == "survival_months") %>%
  select(
    Variable = skim_variable,
    Completeness = complete_rate,
    Mean = numeric.mean,
    SD = numeric.sd,
    Min = numeric.p0,
    Q1 = numeric.p25,
    Median = numeric.p50,
    Q3 = numeric.p75,
    Max = numeric.p100
  )

# Format using GT
surv_summary_gt <- surv_summary %>%
  gt() %>%
  tab_header(title = md("**Table 2: Summary of Survival Months**")) %>%
  fmt_number(columns = where(is.numeric), decimals = 2) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    table_body.hlines.color = "gray90",
    table.border.top.color = "gray80",
    table.border.bottom.color = "gray80",
    row.striping.include_table_body = TRUE,
    data_row.padding = px(4),
    table.align = "center"
  ) %>%
  opt_row_striping() %>%
  tab_source_note(md("*Note:* Summary based on numeric variable `survival_months`."))

# Display the table 
surv_summary_gt

# Save GT object and PNG
saveRDS(surv_summary_gt, here("results", "edatables", "edasurvivalmonths_table_gt.rds"))

# Ensure image folder exists
dir_create(here("results", "edafigures"))

# Save as PNG
gtsave(surv_summary_gt,
       filename = here("results", "edafigures", "edasurvivalmonths_table.png"),
       vwidth = 1000, vheight = 600, zoom = 2)
```

```{r}
# Create a structured summary table
var_summary <- mydata %>%
  summarise_all(class) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Type") %>%
  mutate(Unique_Values = sapply(mydata, function(x) length(unique(x))),
         Missing_Values = sapply(mydata, function(x) sum(is.na(x))))

@@

# Save the structured table in a readable format
summarytable_file = here("results", "edatables", "edasummarytable2.rds")
saveRDS(var_summary, file = summarytable_file)
```
```{r}
# Create publication-ready table
var_summary_gt <- var_summary %>%
  gt() %>%
  tab_header(
    title = md("**Table 1: Variable Summary**")
  ) %>%
  cols_label(
    Variable = "Variable Name",
    Type = "Data Type",
    Unique_Values = "Unique Values",
    Missing_Values = "Missing Values"
  ) %>%
  fmt_number(
    columns = vars(Unique_Values, Missing_Values),
    decimals = 0
  ) %>%
  tab_options(
    table.font.size = "small",
    heading.title.font.size = 12,
    data_row.padding = px(4),
    table.align = "center"
  )

# Display the table
var_summary_gt

# Save the GT table object to RDS
summarytable_file_gt <- here("results", "edatables", "edasummarytable2_gt.rds")
# Save table as GT object (already done)
saveRDS(var_summary_gt, file = here("results", "edatables", "edasummarytable2_gt.rds"))

# ALSO save as high-resolution PNG
gtsave(var_summary_gt,
       filename = here("results", "edafigures", "summarytable.png"),
       vwidth = 1200, vheight = 800, zoom = 2)

@@

```
```{r}
# Generate summary Publication ready table for age_group
p1 <- mydata %>%
  group_by(age_group) %>%
  summarise(Count = n(), .groups = "drop")

# Style it using gt
age_summary_gt <- p1 %>%
  gt() %>%
  tab_header(
    title = md("**Table 3: Distribution of Age Groups**")
  ) %>%
  cols_label(
    age_group = "Age Group",
    Count = "Number of Patients"
  ) %>%
  fmt_number(columns = Count, decimals = 0) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  opt_row_striping() %>%
  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    data_row.padding = px(4),
    table.align = "center",
    table_body.hlines.color = "gray90",
    table.border.top.color = "gray80",
    table.border.bottom.color = "gray80"
  ) %>%
  tab_source_note(md("*Note:* This table summarizes the distribution of patients by age group."))

# Show in EDA HTML output
age_summary_gt

# Save raw table
saveRDS(p1, here("results", "edatables", "edaagesummary.rds"))

# Save styled table
saveRDS(age_summary_gt, here("results", "edatables", "edaagesummary_gt.rds"))

# Ensure folder exists and save PNG
dir_create(here("results", "edafigures"))

gtsave(age_summary_gt,
       filename = here("results", "edafigures", "edaagesummary.png"),
       vwidth = 800, vheight = 500, zoom = 2)
```

Now lets look at the bar plot for age distribution

@@

```
```{r}
#Publication ready table 
# Generate clean summary of SEER Combined Mets at Diagnosis
# Clean and relabel the metastasis site names
p7 <- mydata %>%
  summarise(
    Bone_Mets  = sum(mets_bone == "Yes", na.rm = TRUE),
    Brain_Mets = sum(mets_brain == "Yes", na.rm = TRUE),
    Liver_Mets = sum(mets_liver == "Yes", na.rm = TRUE),
    Lung_Mets  = sum(mets_lung == "Yes", na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Metastasis_Site", values_to = "Count") %>%
  mutate(Metastasis_Site = case_when(
    Metastasis_Site == "Bone_Mets"  ~ "Bone Metastasis",
    Metastasis_Site == "Brain_Mets" ~ "Brain Metastasis",
    Metastasis_Site == "Liver_Mets" ~ "Liver Metastasis",
    Metastasis_Site == "Lung_Mets"  ~ "Lung Metastasis"
  ))

# Format the GT table
mets_summary_gt <- p7 %>%
  gt() %>%
  tab_header(title = md("**Table 4: Metastasis at Diagnosis by Site**")) %>%
  cols_label(
    Metastasis_Site = "Site",
    Count = "Number of Patients"
  ) %>%
  fmt_number(columns = Count, decimals = 0) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
  opt_row_striping() %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    data_row.padding = px(4),
    table.align = "center",
    table_body.hlines.color = "gray90",
    table.border.top.color = "gray80",
    table.border.bottom.color = "gray80"
  ) %>%
  tab_source_note(md("*Note:* This table summarizes the number of patients with bone, brain, liver, or lung metastases at diagnosis."))

# Show in EDA output
mets_summary_gt

# Save RDS versions
saveRDS(p7, here("results", "edatables", "edametsatsummary.rds"))
saveRDS(mets_summary_gt, here("results", "edatables", "edametsatsummary_gt.rds"))

# Save PNG version for manuscript
dir_create(here("results", "edafigures"))

gtsave(mets_summary_gt,
       filename = here("results", "edafigures", "edametsatsummary.png"),
       vwidth = 900, vheight = 600, zoom = 2)
```

### 2025-04-18 code/processing-code/processingcodemz.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

## Renaming all the variables for the ease of analysis 
```{r}
d8 <- d7 %>% dplyr::rename(
    # Demographic Information
    age_group = `Age recode with <1 year olds`,  # Age category of the patient
    sex = Sex,  # Patient's sex
    race = `Race recode (W, B, AI, API)`,  # Race classification
    ethnicity = `Origin recode NHIA (Hispanic, Non-Hisp)`,  # Ethnicity classification
    race_ethnicity = `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`,  # Combined race & ethnicity

>>>>

## Renaming all the variables for the ease of analysis

```{r}
d8 <- d7 %>% dplyr::rename(
    # Demographic Information
    age_group = `Age recode with <1 year olds`,  # Age category of the patient
    sex = Sex,  # Patient's sex
    race = `Race recode (W, B, AI, API)`,  # Race classification
    ethnicity = `Origin recode NHIA (Hispanic, Non-Hisp)`,  # Ethnicity classification
    race_ethnicity = `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`,  # Combined race & ethnicity
